### 2022-06-22 paper.Rmd
editing abstract and introduction

    Anticlustering refers to a collection of methods that partition data sets into disjoint groups while aiming for high between-group similarity or high within-group heterogeneity. Anticlustering is usually approached by maximzing instead of minimizing a clustering objective such as the sum of squared errors used in k-means clustering. Introducing k-plus, this paper presents an extension of the k-means criterion to maximize between-group similarity. While k-means anticlustering minimizes between-group-differences only with regard to the mean of attribute values, k-plus anticlustering can be used to minimize differences with regard to multiple distribution characteristics: the variance, higher order moments such as skewness and kurtosis, and even to preserve covariance structure. Thus, k-plus corresponds to a family of new anticlustering criteria that can be combined in arbitrary ways. It is shown that all k-plus variations can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple criteria with little sacrifice to each single criterion. The k-plus extension is therefore preferred over classical k-means anticlustering when optimizing between-group similarity. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering is a collection of methods to partition elements into disjoint groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: some anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@feo1990]. Anticlustering thereby reverses the logic of its better known twin cluster analysis, which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis---including as hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012, for an overview]---anticlustering is usually formalized as a partitioning process that optimizes an objective function, representing between-group similarity and/or within-group heterogeneity [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split tests into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished on the basis of the input data and the objective function that is used to quantify between-group similarity or within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Maximizing the diversity optimizes within-group heterogeneity, which simultaneously leads to similar groups---if all groups have the same size. Another important criterion based on pairwise dissimilarities is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to find efficient solutions according to both criteria.

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data, using an appropriate clustering objective that is computed on the basis of the features directly.  K-means is probably the best-known clustering method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. The k-plus objective is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics and not only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Hence, the k-plus criterion is not a single criterion for anticlustering, but instead a familiy of objectives implementing different requirements with regard to between-group similarity. After mathematically introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two pratical are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

This section demonstrates how researchers and other data analysists can easily employ k-plus anticlustering using the R package `anticlust` [@papenberg2020], using an interface function called `kplus_anticlustering()`, which is available from version 0.6.3 onward. The package can be installed in the R environment using the `install.packages()` command:

>>>>

    Anticlustering is a collection of methods to partition data sets into disjoint groups while aiming for high between-group similarity or high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximzing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. While k-means anticlustering minimizes between-group differences only with regard to mean attribute values, k-plus anticlustering can be used to minimize differences with regard to several additional distribution characteristics, including the variance, skewness and kurtosis, and even covariance structure. It is shown that all k-plus variations can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple criteria. In particular, optimizing similarity with regard to variances as well as means comes with almost no cost to either criterion; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Higher order moments and covariance structure can also succesfully be equalized, while fulfilling multiple criteria at the same time is improved when the sample size increases. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering is a collection of methods to partition elements into disjoint groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: some anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@feo1990]. Anticlustering thereby reverses the logic of its better known twin cluster analysis, which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis---including hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012, for an overview]---anticlustering is usually formalized as a partitioning process that optimizes an objective function that represents between-group similarity and/or within-group heterogeneity [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology [@steinley2006]. For example, anticlustering can be used to split tests into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished on the basis of the input data and the objective function that is used to quantify between-group similarity or within-group heterogeneity. Some anticlustering objectives are based on pairwise dissimilarity ratings such as the Euclidean distance. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019]. By summing dissimilarities within-groups, the diversity is primarily a measure of within-group heterogeneity. For equal-sized groups, however, maximizing the diversity simultaneously minimizes the average dissimilarity between elements in different groups, thereby maximizing between-group similarity.

Another important anticlustering criterion that is based on pairwise dissimilarities is the *dispersion*, i.e., the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

[^averagesum]:

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the elements. A different anticlusting approach is based on clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known clustering method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. The k-plus objective is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Hence, the k-plus criterion is not a single criterion for anticlustering, but a familiy of objectives implementing different requirements with regard to between-group similarity. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two pratical are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

This section demonstrates how readers can easily employ k-plus anticlustering using the R package `anticlust` [@papenberg2020], using an interface function called `kplus_anticlustering()`, which is available from version 0.6.3 onward. The package can be installed in the R environment using the `install.packages()` command:
